Experience Report: Haskell in Computational Biology

### 2012-06-28 everygen.tex
wordsmithing, death of age, revised search

everyPt :: SearchGen pt r -> Age -> Scored pt
         -> Rand r [Aged (Utility (Scored pt))]
everyPt ss age startPt = do
  successors <- mapM (nextPt ss) (repeat startPt)
  tagged <- zipWithM agedUtility successors [succ age..]
  let (useless, Aged (Useful newPt) newAge : _) =
                        span (isUseless . unAged) tagged
  nextPts <- everyPt ss newAge newPt
  return $ Aged (Useful startPt) age : useless ++ nextPts

  where agedUtility pt age =
           utility ss move >>= \u -> return $ Aged u age
         where move = Move { older = startPt, younger = pt
                           , youngerAge = age }

>>>>

everyPt :: RandomGen r
        => SearchGen pt r -> CCost -> Scored pt
        -> Rand r [CCosted (Utility (Scored pt))]
everyPt sg age startPt = do
  successors <- mapM (nextPt sg) (repeat startPt)
  tagged <- zipWithM agedUtility successors [succ age..]
  let (useless, CCosted (Useful newPt) newCCost : _) =
                        span (isUseless . unCCosted) tagged
  nextPts <- everyPt sg newCCost newPt
  return $ CCosted (Useful startPt) age : useless ++ nextPts

  where agedUtility pt age =
           utility sg move >>= \u -> return $ CCosted u age
         where move = Move { older = startPt, younger = pt
                           , youngerCCost = age }

### 2012-06-28 mrfy_experience_report.tex
wordsmithing, death of age, revised search

% \category{CR-number}{subcategory}{third-level}
% 
% \terms
% term1, term2
% 
% \keywords
% keyword1, keyword2

@@

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds $e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum.
The sum of an~$a'$~term and a $V'_{j-1}(i-1)$ term 
is computed by function \texttt{avSum},
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively.
\smallverbatiminput{vfix}
The entire right-hand side is wrapped in \texttt{extend~Mat}.
This call performs a computation \emph{not} shown in \figref{viterbi}:
in~addition to computing the most likely BUZZWORD,
\mrfy\ also computes most likely path (sequence of states).
Function \texttt{extend} adds~$M$ to the head of that path.

@@

To~account for beta strands, computational biologists must use both a
richer model of protein structure and a revised version of Viterbi's
equations.
The~richer model recognizes that only certain pairs of amino acids are
able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on~$x_{\pairedwith j}$.
The~relevant term in
Viterbi's equations 
$V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
Columns~$j$ and $\pairedwith j$ can be as close together as a few
columns or as far apart as a few hundreds of columns.
Because $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
also on the nonlocal $x_{\pairedwith j}$, 
dynamic programming cannot compute the maximum likelihood quickly
\cite{Menke:2010ti,Daniels:2012}.
MRFy explores new methods of \emph{approximating}
maximum likelihood.

The presence of beta strands also changes the model.
Within a beta strand, amino acids are not inserted or deleted;
a~bonded pair of beta strands is modeled by
 a pair of sequences of match states.
Between the beta strands, the model is structured as before.
The~combined model, an~example of which is shown in \figref{mrf}, is called a
\textit{Markov random field}.

@@

MRFy searches for placements stochastically,
trying to minimize scores.
We~are exploring several stochastic algorithms: random-mutation hill
climbing, simulated  
annealing, multistart simulated annealing, and a genetic algorithm.
Our~exploration exploits higher-order functions, existentially
quantified types, and (to a degree) lazy evaluation.

\begin{figure}
\ifpdfmadness
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
\else
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.eps}} 
\fi
In this Markov random field,
each shaded node represents a beta-strand position.
Nodes connected by dashed edges are hydrogen-bonded.

@@

We~describe \mrfy's search abstractly: \mrfy\ computes a sequence
of \emph{points}. 
Each point contains
a collection of {placements}; different searches represent
collections differently.
Each point also has a \texttt{Score}; \mrfy\ looks for points with
good scores.

Ideally, MRFy's search would use now-classic lazy, modular
technique advocated by \citet{hughes:why}, in which one function
computes an infinite sequence of points, and another function
uses a finite prefix to decide on an approximation.
But because MRFy's search is stochastic, 
making MRFy's search lazy and modular is not so easy.

To illustrate the difficulties, we~use our simplest search:
random hill climbing.
The space of beta-strand placements has no notion of slope or
gradient,
so at any given point in the search space, this search moves a random
distance in a random 
direction.
If~it leads to a better point, the move
is \emph{useful};
otherwise it is \emph{useless}.
\smallverbatiminput{utility}
(We~also use \texttt{Useful} and \texttt{Useless} to tag \emph{points}.)
With luck, an~infinite sequence of useful moves converges
at a local optimum.

\mrfy's search path follows only useful moves;
if a move proves useless, 
\mrfy\ abandons it and tries again (with a new random seed) at the
previous point.
Ideally, \mrfy\ would search by composing a \emph{generator}
that produces an infinite sequence of moves, 
a \emph{filter} that selects the useful moves, and 
a \emph{test} function that enumerates finitely many useful moves
and returns the final destination.
But the generator may produce an infinite sequence of useless moves.
(For example, if \mrfy\ should find a global optimum, every move from
that point is guaranteed to be useless.)
Given an infinite sequence of useless inputs, a filter would not be
productive, and a test function would diverge.

We~address this problem by combining ``generate and filter'' into a
single abstraction, which has type \texttt{SearchGen pt r}.
Type variable~\texttt{pt} is a point in the search
space, and type variable~\texttt{r} is a
random-number generator:
\smallverbatiminput{gen.tex}
The monadic computation \texttt{pt0} randomly selects a starting point
for search;
\texttt{nextPt} produces a new point from
an existing point.
Because scoring may be expensive, both \texttt{pt0}
and \texttt{nextPt} use \emph{scored} points, and they can
reuse scores from previous points.

Why do we have \texttt{pt0} and \texttt{nextPt}
instead 
of a single function with type \mbox{\texttt{Rand r [Scored pt]}},
which could produce an infinite list?
Because after the first move
the starting point for the next move depends on whether the first move
was useful, whereas the first move is always useful.
That decision is made by the \texttt{utility} function,
which scrutinizes a move represented as follows:
\smallverbatiminput{move}
The decision about utility uses not only a source of randomness but
also the \emph{age} of the search, which corresponds to the
computational cost to reach a point in the search.
For example, in~simulated annealing, as the search progresses,
the \texttt{utility} function becomes less likely to express a move
that results in a worse score.

>>>>

\category{J.3}{Biology and genetics}{}
\category{D.3.2}{Language Classifications}{Applicative (functional) languages}
% \category{CR-number}{subcategory}{third-level}
% 
% \terms
% term1, term2
% 
% \keywords
% keyword1, keyword2

@@

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum.
The sum of an~$a'$~term and a $V'_{j-1}(i-1)$ term 
is computed by function \texttt{avSum},
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively.
\smallverbatiminput{vfix}
The entire right-hand side is wrapped in \texttt{extend~Mat}.
This call performs a computation \emph{not} shown in \figref{viterbi}:
in~addition to computing the most likely BUZZWORD,
\mrfy\ also computes most likely path (sequence of states).
Function \texttt{extend} adds~$M$ to the head of that path.

@@

To~account for beta strands, computational biologists add
a term to Viterbi's equations, and they build
richer models of protein structure.
%%  The~richer model recognizes that only certain pairs of amino acids are
%%  able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on~$x_{\pairedwith j}$.
The~relevant term in
Viterbi's equations 
$V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
Columns~$j$ and $\pairedwith j$ can be as close together as a few
columns or as far apart as a few hundreds of columns.
Because $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
also on the nonlocal $x_{\pairedwith j}$, 
dynamic programming cannot compute the maximum likelihood quickly
\cite{Menke:2010ti,Daniels:2012}.

The new terms in Viterbi's equations are accompanied by a new model.
Within a beta strand, amino acids are not inserted or deleted, so
a bonded pair of beta strands is modeled by
 a pair of sequences of match states.
Between the beta strands, the model is structured as before.
The~combined model, an~example of which is shown in \figref{mrf}, is called a
\textit{Markov random field}.

@@

MRFy searches for good placements stochastically.
\mrfy\ implements
random hill
climbing, simulated  
annealing, multistart simulated annealing, and a genetic algorithm.
These algorithms share much code, and they use higher-order functions,
existentially 
quantified types, and lazy evaluation.

We~describe \mrfy's search abstractly: \mrfy\ computes a sequence
of \emph{points} in a search space.
A~point is typically a single placement or perhaps a collection of
placements. 
Each point also has a \texttt{Score}; \mrfy\ looks for points with
good scores.

Ideally, MRFy would use now-classic lazy, modular
technique advocated by \citet{hughes:why}, in which one function
computes an infinite sequence of points, and another function
uses a finite prefix to decide on an approximation.
But because MRFy's search is stochastic, 
making MRFy's search lazy and modular is not so easy.

To illustrate the difficulties, we~discuss our simplest search:
random hill climbing.
%The space of beta-strand placements has no notion of slope or
%gradient,
From any given point in the search space, this search moves a random
distance in a random 
direction.
If~the move leads to a better point, it
is \emph{useful};
otherwise it is \emph{useless}.
\smallverbatiminput{utility}
(We~also use \texttt{Useful} and \texttt{Useless} to tag \emph{points}.)
With luck, an~infinite sequence of useful moves converges
at a local optimum.

\mrfy's search path follows only useful moves;
if a move is useless, 
\mrfy\ abandons it and moves again (in a new random direction) from the
previous point.
Ideally, \mrfy\ would search by composing a \emph{generator}
that produces an infinite sequence of moves, 
a \emph{filter} that selects the useful moves, and 
a \emph{test} function that enumerates finitely many useful moves
and returns the final destination.
But a generator may produce an infinite sequence of useless moves.
(For example, if \mrfy\ should find a global optimum, every move from
that point is guaranteed to be useless.)
Given an infinite sequence of useless inputs, a filter would not
produce any values, and the search would diverge.

\begin{figure}
\ifpdfmadness
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
\else
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.eps}} 
\fi
Each shaded node represents a beta-strand position.
Nodes connected by dashed edges are hydrogen-bonded.

@@

We~address this problem by combining ``generate and filter'' into a
single abstraction, which has type \texttt{SearchGen pt r}.
Type variable~\texttt{pt} is a point in the search
space, and \texttt{r} is a
random-number generator.
\texttt{Rand~r} is a monad of stochastic computations:
\smallverbatiminput{gen.tex}
The monadic computation \texttt{pt0} randomly selects a starting point
for search;
\texttt{nextPt} produces a new point from
an existing point.
Because scoring can be expensive, both \texttt{pt0}
and \texttt{nextPt} use \emph{scored} points, and they can
reuse scores from previous points.

Why have \texttt{pt0} and \texttt{nextPt}
instead 
of a single function with type \mbox{\texttt{Rand r [Scored pt]}},
which could produce an infinite list of points?
Because if the infinite list contains a useless point,

the starting point for the next move depends on whether the first move
was useful, whereas the first move is always useful.
That decision is made by the \texttt{utility} function,
which scrutinizes a move represented as follows:
\smallverbatiminput{move}
The decision about utility uses not only a source of randomness but
also the \emph{age} of the search, which corresponds to the
computational cost to reach a point in the search.
For example, in~simulated annealing, as the search progresses,
the \texttt{utility} function becomes less likely to express a move
that results in a worse score.
